---
title: "Methylation_Mark_Analysis"
author: "HM Putnam"
date: "2024-08-26"
output: html_document
---

```{r}
library(tidyverse)
library(ggsci)
```

5 methyl cytosine 5mC Quantitifcation
Kit: Methylflash Methylated DNA quantificiation Kit (Fluorometric) (Reference: P-1035)								https://www.epigentek.com/docs/P-1035.pdf

Relative Quantification: To determine the relative methylation status of two different DNA samples,
simple calculation for percentage of 5-mC in total DNA can be carried out using the following formula:

Relative %5mC = ((Sample RFU - MF3 RFU)/ input DNA ng) / ((MF4 RFU - MF3 RFU)x 2 /amount of input positive control (MF4) in ng)

S is the amount of input sample DNA in ng.
P is the amount of input positive control (MF4) in ng.

Acropora hyacinthus
```{r}
ahya <- read.csv("RAnalysis/data/Ahyacinthus_5mC.csv",  na.strings = "NA")

str(ahya)
as.numeric

stnds <- ahya %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

ahya <- ahya %>% filter(Species=="A. hyacinthus") 

MF3.RFU.ahya <- stnds$mean.RFU[[1]]
MF4.RFU.ahya <- stnds$mean.RFU[[2]]
MF4.ng.ahya <- 10

ahya$rel.percent.5mC <- (((ahya$Fluoro.530.590 -MF3.RFU.ahya )/ ahya$DNA.ng) / ((MF4.RFU.ahya - MF3.RFU.ahya)* 2 /MF4.ng.ahya))*100

ahya.27 <- ahya %>%
filter(Temperature ==27)

```

Acropora cytherea
```{r}
acyth <- read.csv("RAnalysis/data/Acytherea_5mC.csv",  na.strings = "NA")

str(acyth)
as.numeric

stnds <- acyth %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

acyth <- acyth %>% filter(Species=="A. cytherea") 

MF3.RFU.acyth <- stnds$mean.RFU[[1]]
MF4.RFU.acyth <- stnds$mean.RFU[[2]]
MF4.ng.acyth <- 10

acyth$rel.percent.5mC <- (((acyth$Fluoro.530.590 -MF3.RFU.acyth )/ acyth$DNA.ng) / ((MF4.RFU.acyth - MF3.RFU.acyth)* 2 /MF4.ng.acyth))*100

acyth <- acyth %>%
filter(Temperature ==27)

```
Acropora retusa
```{r}
aret <- read.csv("RAnalysis/data/Aretusa_5mC.csv",  na.strings = "NA")

str(aret)
as.numeric

stnds <- aret %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

aret <- aret %>% filter(Species=="A. retusa") 

MF3.RFU.aret <- stnds$mean.RFU[[1]]
MF4.RFU.aret <- stnds$mean.RFU[[2]]
MF4.ng.aret <- 10

aret$rel.percent.5mC <- (((aret$Fluoro.530.590 -MF3.RFU.aret )/ aret$DNA.ng) / ((MF4.RFU.aret - MF3.RFU.aret)* 2 /MF4.ng.aret))*100

aret <- aret %>%
filter(Temperature ==27) %>%
filter(!Stage == "Stage 12")
```
Acropora pulchra
```{r}
apul <- read.csv("RAnalysis/data/Apulchra_5mC.csv",  na.strings = "NA")

str(apul)

stnds <- apul %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

apul <- apul %>% filter(Species=="A. pulchra") 

MF3.RFU.apul <- stnds$mean.RFU[[1]]
MF4.RFU.apul <- stnds$mean.RFU[[2]]
MF4.ng.apul <- 10

apul$rel.percent.5mC <- (((apul$Fluoro.530.590 -MF3.RFU.apul )/ apul$DNA.ng) / ((MF4.RFU.apul - MF3.RFU.apul)* 2 /MF4.ng.apul))*100

apul <- apul %>%
filter(Temperature ==27)

```

```{r}
#examine only 27

all.species.5mC <- rbind(ahya.27, acyth, aret,  apul)
unique(all.species.5mC$Stage)

devo.27 <- all.species.5mC %>% 
  mutate(Stage = fct_relevel(Stage, "Sperm",      "Oocytes",   "Stage 1 ",   "Stage 2",    
                             "Stage 3",    "Stage 4",    "Stage 5", "Stage 5.5",    "Stage 6",   "Stage 7",    
                             "Stage 7.5", "Stage 8",    "Stage 9",    "Stage 10",   "Stage 11",   "Stage 12",   
                             "Stage 13",   "Stage 14",  "Stage 15",   "Stage 16",   "Stage 17",   "Stage 17.5"))

summary_stats.devo.27 <- devo.27 %>%
  group_by(Species, Stage) %>%
  summarize(
    mean_5mC = mean(rel.percent.5mC, na.rm = TRUE),
    sem_5mC = sd(rel.percent.5mC, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

summary_stats.devo.27 %>%
  ggplot(aes(x = Stage, y = mean_5mC, group = Species, color = Species)) +
  geom_line(aes(linetype = Species), size = 1) +  # Line connecting mean points
  geom_point(size = 3) +  # Mean points
  geom_errorbar(aes(ymin = mean_5mC - sem_5mC, ymax = mean_5mC + sem_5mC), width = 0.2) +  # SEM bars
  labs(
    x = "Stage",
    y = "Mean Relative % of 5mC"
  ) +
  scale_color_npg()+
  scale_fill_npg()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability


```

Reduce to commonly sampled timepoints and analyze with ANOVA
```{r}

devo.27.stages <- devo.27 %>% 
  filter(!Stage == "Stage 5.5") %>% 
  filter(!Stage == "Stage 7.5") %>% 
  filter(!Stage == "Stage 12")%>% 
  filter(!Stage == "Stage 17")%>% 
  filter(!Stage == "Stage 17.5")

str(devo.27.stages)
devo.27.stages$Species <- as.factor(devo.27.stages$Species)

unique(devo.27.stages$Stage)

unique(devo.27.stages$Development)

devo.27.stages <- devo.27.stages %>% 
  mutate(Development = fct_relevel(Development, "Sperm", "Oocytes", "Fertilized oocyte", "4 cells", "8 cells", "Morula",          "Prawn Chip", "Blastula", "Late Blastula", "Early Bowl", "Bowl", "Blastulation end", "Gastrula", "Early tear drop",
"Tear drop", "Late tear drop", "Early elongated"))

stage.5mC.res <- aov(log10(rel.percent.5mC) ~ Stage*Species, data= devo.27.stages)
summary(stage.5mC.res)
hist(stage.5mC.res$residuals)
boxplot(stage.5mC.res$residuals)
plot(stage.5mC.res$residuals, stage.5mC.res$fitted.values)

Tukey.ph <- TukeyHSD(stage.5mC.res)

summary_stats.devo.27 <- devo.27.stages %>%
  group_by(Species, Stage, Development) %>%
  summarize(
    mean_5mC = mean(rel.percent.5mC, na.rm = TRUE),
    sem_5mC = sd(rel.percent.5mC, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

summary_stats.devo.27 %>%
  ggplot(aes(x = Development, y = mean_5mC, group = Species, color = Species)) +
  geom_line(aes(linetype = Species), size = 1) +  # Line connecting mean points
  geom_point(size = 3) +  # Mean points
  geom_errorbar(aes(ymin = mean_5mC - sem_5mC, ymax = mean_5mC + sem_5mC), width = 0.2) +  # SEM bars
  labs(
    x = "Stage",
    y = "Mean Relative % of 5mC"
  ) +
  scale_color_npg()+
  scale_fill_npg()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability



```



DNA Hydroxymethylation 5hmC Quantification

Kit: Methylflash Hydroxymethylated DNA quantificiation Kit (Fluorometric) (Reference: P-1037)									https://www.epigentek.com/docs/P-1037.pdf


Relative Quantification: To determine the relative hydroxymethylation status of two different DNA
samples, simple calculation for the precentage of 5-hmC in your total DNA can be carried out using the
following formula: 

Relative %5hmC = ((Sample RFU - HF4 RFU)/ input DNA ng) / ((HF5 RFU - HF4 RFU)x 5 /amount of input positive control (HF5) in ng)

S is the amount of input sample DNA in ng.
P is the amount of input positive control (HF5) in ng.


Acropora hyacinthus
```{r}
ahya.5hmC <- read.csv("RAnalysis/data/Ahyacinthus_5hmC.csv",  na.strings = "NA")

str(ahya.5hmC)

stnds.ahya.5hmC <- ahya.5hmC %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

ahya.5hmC <- ahya.5hmC %>% filter(Species=="A. hyacinthus") 

HF4.RFU.ahya.5hmC <- stnds.ahya.5hmC$mean.RFU[[1]]
HF5.RFU.ahya.5hmC <- stnds.ahya.5hmC$mean.RFU[[2]]
HF5.ng.ahya.5hmC <- 10

ahya.5hmC$rel.percent.5mC <- (((ahya.5hmC$Fluoro.530.590 -HF4.RFU.ahya.5hmC )/ ahya.5hmC$DNA.ng) / ((HF5.RFU.ahya.5hmC - HF4.RFU.ahya.5hmC)* 5 /HF5.ng.ahya.5hmC))*100

ahya.5hmC <- ahya.5hmC %>%
filter(Temperature ==27)

```


Acropora cytherea 5hmC
```{r}
acyth.5hmC <- read.csv("RAnalysis/data/Acytherea_5hmC.csv",  na.strings = "NA")

str(acyth.5hmC)

stnds.acyth.5hmC <- acyth.5hmC %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

acyth.5hmC <- acyth.5hmC %>% filter(Species=="A. cytherea") 

HF4.RFU.acyth <- stnds.acyth.5hmC$mean.RFU[[1]]
HF5.RFU.acyth <- stnds.acyth.5hmC$mean.RFU[[2]]
HF5.ng.acyth <- 10

acyth.5hmC$rel.percent.5mC <- (((acyth.5hmC$Fluoro.530.590 -HF4.RFU.acyth )/ acyth.5hmC$DNA.ng) / ((HF5.RFU.acyth - HF4.RFU.acyth)* 5 /HF5.ng.acyth))*100

acyth.5hmC <- acyth.5hmC %>%
filter(Temperature ==27)

```
Acropora pulchra 5hmC
```{r}
apul.5hmC <- read.csv("RAnalysis/data/Apulchra_5hmC.csv",  na.strings = "NA")

str(apul.5hmC)

stnds.apul.5hmC <- apul.5hmC %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

apul.5hmC <- apul.5hmC %>% filter(Species=="A. pulchra") 

HF4.RFU.apul <- stnds.apul.5hmC$mean.RFU[[1]]
HF5.RFU.apul <- stnds.apul.5hmC$mean.RFU[[2]]
HF5.ng.apul <- 10

apul.5hmC$rel.percent.5mC <- (((apul.5hmC$Fluoro.530.590 -HF4.RFU.apul)/ apul.5hmC$DNA.ng) / ((HF5.RFU.apul - HF4.RFU.apul)* 5 /HF5.ng.apul))*100

apul.5hmC <- apul.5hmC %>%
filter(Temperature ==27)

```

Acropora retusa 5hmC
```{r}
aret.5hmC <- read.csv("RAnalysis/data/Aretusa_5hmC.csv",  na.strings = "NA")

str(aret.5hmC)

stnds.aret.5hmC <- aret.5hmC %>% filter(Species=="Standard") %>%
  group_by(Stage) %>%
  summarize(mean.RFU=mean(Fluoro.530.590))

aret.5hmC <- aret.5hmC %>% filter(Species=="A. retusa") 

HF4.RFU.aret <- stnds.aret.5hmC$mean.RFU[[1]]
HF5.RFU.aret <- stnds.aret.5hmC$mean.RFU[[2]]
HF5.ng.aret <- 10

aret.5hmC$rel.percent.5mC <- (((aret.5hmC$Fluoro.530.590 -HF4.RFU.aret )/ aret.5hmC$DNA.ng) / ((HF5.RFU.aret - HF4.RFU.aret)* 5 /HF5.ng.aret))*100

aret.5hmC <- aret.5hmC %>%
filter(Temperature ==27)

```

5hmC in all species at 27°C
```{r}
#examine only 27

all.species.5hmC <- rbind(ahya.5hmC, acyth.5hmC, apul.5hmC, aret.5hmC)
unique(all.species.5hmC$Stage)

devo.5hmC.27 <- all.species.5hmC %>% 
  mutate(Stage = fct_relevel(Stage, "Sperm",      "Oocytes",   "Stage 1 ",   "Stage 2",    
                             "Stage 3",    "Stage 4",    "Stage 5", "Stage 5.5",    "Stage 6",   "Stage 7",    
                             "Stage 7.5", "Stage 8",    "Stage 9",    "Stage 10",   "Stage 11",   "Stage 12",   
                             "Stage 13",   "Stage 14",  "Stage 15",   "Stage 16",   "Stage 17",   "Stage 17.5"))

summary_stats.devo.27.5hmC <- devo.5hmC.27 %>%
  group_by(Species, Stage) %>%
  summarize(
    mean_5mC = mean(rel.percent.5mC, na.rm = TRUE),
    sem_5mC = sd(rel.percent.5mC, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

summary_stats.devo.27.5hmC %>%
  ggplot(aes(x = Stage, y = mean_5mC, group = Species, color = Species)) +
  geom_line(aes(linetype = Species), size = 1) +  # Line connecting mean points
  geom_point(size = 3) +  # Mean points
  geom_errorbar(aes(ymin = mean_5mC - sem_5mC, ymax = mean_5mC + sem_5mC), width = 0.2) +  # SEM bars
  labs(
    x = "Stage",
    y = "Mean Relative % of 5mC"
  ) +
  scale_color_npg()+
  scale_fill_npg()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability


```

Reduce to commonly sampled timepoints and analyze with ANOVA
```{r}

devo.5hmC.27.stages <- devo.5hmC.27 %>% 
  filter(!Stage == "Stage 5.5") %>% 
  filter(!Stage == "Stage 7.5") %>% 
  filter(!Stage == "Stage 12")%>% 
  filter(!Stage == "Stage 17")%>% 
  filter(!Stage == "Stage 17.5")

str(devo.5hmC.27.stages)
devo.5hmC.27.stages$Species <- as.factor(devo.5hmC.27.stages$Species)

unique(devo.5hmC.27.stages$Stage)

unique(devo.5hmC.27.stages$Development)

devo.5hmC.27.stages <- devo.5hmC.27.stages %>% 
  mutate(Development = fct_relevel(Development, "Sperm", "Oocytes", "Fertilized oocyte", "4 cells", "8 cells", "Morula",          "Prawn Chip", "Blastula", "Late Blastula", "Early Bowl", "Bowl", "Blastulation end", "Gastrula", "Early tear drop",
"Tear drop", "Late tear drop", "Early elongated"))

stage.5hmC.res <- aov(log10(rel.percent.5mC) ~ Stage*Species, data= devo.5hmC.27.stages)
summary(stage.5hmC.res)
hist(stage.5hmC.res$residuals)
boxplot(stage.5hmC.res$residuals)
plot(stage.5hmC.res$residuals, stage.5hmC.res$fitted.values)

Tukey.ph.5hmC <- TukeyHSD(stage.5hmC.res)

summary_stats.5hmC.27 <- devo.5hmC.27.stages %>%
  group_by(Species, Stage, Development) %>%
  summarize(
    mean_5mC = mean(rel.percent.5mC, na.rm = TRUE),
    sem_5mC = sd(rel.percent.5mC, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

summary_stats.5hmC.27 %>%
  ggplot(aes(x = Development, y = mean_5mC, group = Species, color = Species)) +
  geom_line(aes(linetype = Species), size = 1) +  # Line connecting mean points
  geom_point(size = 3) +  # Mean points
  geom_errorbar(aes(ymin = mean_5mC - sem_5mC, ymax = mean_5mC + sem_5mC), width = 0.2) +  # SEM bars
  labs(
    x = "Stage",
    y = "Mean Relative % of 5hmC"
  ) +
  scale_color_npg()+
  scale_fill_npg()+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability



```